Radio Crickets: Chirping Jets from Black Hole Binaries Entering their Gravitational Wave Inspiral

### 2015-07-21 Paper/paper.tex
Correct typos

\begin{abstract}
  We study a novel electromagnetic signature of supermassive black
  hole binaries whose inspiral stars being dominated by gravitational
  wave (GW) emission. Combined GW and electromagnetic observations of
  compact binary mergers would allow to use them as standard sirens
  for cosmology and a testing ground of general relativity.  If one of
  the binary members produces a radio jet, the helical chirping
  morphology of the jet can be used to infer the binary parameters.
  For binaries with mass $10^7$--$10^{10}$ M$_\odot$ at redshifts
  $z<0.5$, the jet precesses along a biconical surface due to the
  binary's orbital motion.  When the binary enters the GW phase of its
  evolution, the opening angle widens, the jet exhibits milliarcsecond
  scale wiggles and the conical surface of jet precession is twisted
  due to the superluminal motion.  Monitoring these features in
  current and archival data will place a lower limit on sources that
  could be detected by eLISA and Pulsar Timing Arrays.  In the future,
  microarcsecond interferometry with the Square Kilometer Array will
  increase the potential usefulness of this technique.
\end{abstract}

@@

Binaries of supermassive black holes (BHs) arise naturally as a result
of mergers of galaxies in the context of the heirarchical galaxy
formation \citep{2000MNRAS.311..576K, 2002MNRAS.336L..61H,
  2008ApJ...676...33D, 2012MNRAS.422.1306K}.  After a galaxy merger,
infalling black holes lose their angular momentum in three stages
before coalescence \citep{1980Natur.287..307B, 2005LRR.....8....8M,
  2011ASL.....4..181C}.  In the first stage, the black holes sink to
the center of the gravitational potential of the merger remnant due to
dynamical friction and form a gravitationally bound binary.  The newly
formed binary then continues to lose its angular momentum and shrink
by scattering ambient gas and stars.  In the final stage of the
binary's evolution, emission of gravitational waves (GWs) becomes the
predominant mode angular momentum loss, which results in the
coalescence of the black holes.

General relativisitic (GR) numerical simulations can predict the
binary's evolution in its final stage up to coalescence
\citep{2005PhRvL..95l1101P, 2006PhRvL..96k1102B, 2006PhRvL..96k1101C}.
This has stimulated interest in predicting related electromagnetic
signals of black hole coalescence.  Effects considered in the
literature include afterglows due to infall of gas on the remnant
black hole \citep{2005ApJ...622L..93M}, electromagnetic variability in
the circumbinary disk due to shocks induced by the sudden mass loss of
the binary due to GW emission at coalescence
\citep{2007APS..APR.S1010B}, flares from shocked remnants of accretion
disks around recoiled remnant black hole \citep{2007PhRvL..99d1103L,
  2008ApJ...682..758S, 2009CQGra..26i4032H}, quasi-periodic
variability in gas accretion in the early stages of black hole merger
\citep{2008ApJ...672...83M, 2009CQGra..26i4032H, 2012MNRAS.427.2680K},
evolution in the profiles of broad emission lines shortly before
merger \citep{2013MNRAS.432.1468M}, characteristic evolution in
thermal emission due to circumbinary accretion
\citep{2015MNRAS.446L..36F}, brightening of accretion disk due to
viscous dissipation of gravitational waves
\citep{2008PhRvL.101d1101K}, and prompt tidal disruption of stars by
the recoiled remnant \citep{2011MNRAS.412...75S}.

In addition to testing GR in the strong field limit, observations of
electromagnetic counterparts will help localize sources of GWs for
future detectors such as the Evolved Laser Interferometer Space
Antenna (eLISA)\footnote{\url{https://www.elisascience.org}}, which would be
capable of observing the peak GW luminosity for BH coalescence of
$\sim 10^{57}$ erg$/$s up to large cosmological distances
\citep{2003CQGra..20S..65H, 2013CQGra..30x4009S}.  Simultaneous
detection of GW and electromagnetic signals from coalescing binaries
will yield rich scientific pay-offs \citep{2003CQGra..20S..65H,
  2005ApJ...629...15H}.  For example, by a redshift measurement of
eLISA sources could constraint cosmological parameters.  Measurement
of the black hole masses and spins from the GW signal will constrain
accretion physics during coalescence.  But even before eLISA is
launched, a search of electromagnetic signatures may point to the
expected abundance of GW sources.

In this Letter we study the morphology of a jet emitted by one of the
binary members during its final stage of evolution before coalescence,
i.e., when the binary's dynamics is dominated by GW emission.  We show
that the helical chirping morphology of the jet can be used to infer
the binary parameters.

@@

We study a BH binary formed following a gas-rich merger of two
galaxies as a result of the processes described above.  We consider
black holes with masses $M_1$ and $M_2$ in a circular orbit of radius
$a$ around each other.  Let $M=M_1+M_2$.  For simplicity, we consider
an equal-mass binary ($M_1=M_2$) on a circular orbit. We will consider
unequal black holes masses and elliptical orbits below.  The orbital
speed of each black hole is given by \citep{2010PhRvD..81d7503L}
\begin{equation}
  v = \frac{1}{2}\left(\frac{GM}{a}\right)^{1/2}\!\!\!\! = 5.8\times 10^3\, \mathrm{km~s}^{-1}M_8^{1/2}a_{16}^{-1/2},
  \label{eqn:vorb}
\end{equation}
where $a_{16}\equiv (a/10^{16}$ cm$)$ and $M_8\equiv (M/10^8$
M$_\odot)$ are the binary separation and mass,
respectively.\footnote{Note that $10^{16} \mathrm{cm}\approx 3.4\times
  10^{-3} \mathrm{pc}$.  The Schwarzchild radius of a $10^8$~M$_\odot$
  black hole is $9.6\times 10^{-6}$~pc, or $\sim 3\times 10^{13}$~cm.}
The orbital period is
\begin{equation}
  P = 2\pi\left(\frac{GM}{a^3}\right)^{-1/2}\!\!\!\! = 1.72\, \mathrm{yr}\, a_{16}^{3/2} M_8^{-1/2}.
\end{equation}

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \draw (0,0) ellipse (3 and 0.5);
      \fill[black] (0,0) circle (1pt);
      \draw [-angle 45,thick] (2,-0.39) -- (3.5,-0.1);
      \draw [-angle 45,thick,dashed,dash pattern = on 4pt off 1pt] (2,-0.39) -- (2,1.5);
      \draw [-angle 45,thick] (2,-0.39) -- (2.5,0.8);
      \draw [-angle 45,thick,red] (2,-0.39) -- (3.5,0.8);
      \fill[black] (-2,0.39) circle (2pt);
      \fill[black] (2,-0.39) circle (2pt);
      \draw (4.1,-0.1) node {\large ${\mathbf{v}}_\mathrm{orb}$};
      \draw (2.7,1.0) node {\large ${\mathbf{v}}_\mathrm{jet}$};
      \draw (4.0,1.0) node {\large $\tilde{\mathbf{v}}_\mathrm{jet}$};
      \draw (2.0,1.8) node {\large $\mathbf{L}_\mathrm{orb}$};
    \end{tikzpicture}
  \end{center}
  \caption{Jet and orbital configuration. The orbital velocity
    $\mathbf{v}_\mathrm{orb}$ causes the jet to precess as the BH
    moves along the orbit. The net jet velocity is
    $\tilde{\mathbf{v}}_\mathrm{jet}={\mathbf{v}}_\mathrm{orb}+{\mathbf{v}}_\mathrm{jet}$,
    where $\mathbf{v}_\mathrm{jet}$ is its ejection velocity at the
    source.}
  \label{fig:BH_orbit}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \draw [-angle 45,dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (3.0,0);
      \draw [-angle 45,dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (0,4.0);
      \draw [-angle 45,thick] (0,0) -- (2.0,4.0);
      \draw [-angle 45,thick,red] (0,0) -- (5.0,4.0);
      \draw (1.0,2.0) arc (63:90:2.2);
      \draw (2.0,1.6) arc (38.6:63:2.56);
      \draw (0.5,2.5) node {\large $\chi$};
      \draw (1.90,2.3) node {\large $\psi$};
      \draw [dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (3.0,6.0);
      \draw (0.0,4.5) node {\large $\mathbf{L}_\mathrm{orb}$};
      \draw (3.5,0.0) node {\large $\mathbf{v}_\mathrm{orb}$};
      \draw (1.0,3.0) node {\large $\mathbf{v}_\mathrm{jet}$};
      \draw (5.3,4.2) node {\large \textcolor{red}{$\mathbf{\tilde v}_\mathrm{jet}$}};
    \end{tikzpicture}
  \end{center}
  \caption{The orbital velocity $\mathbf{v}_\mathrm{orbital}$
    introduces the jet precision angle $\psi$.  Without the orbital
    motion, $\psi=0$ and the jet traces a cylindrical surface.  With
    the orbital motion, the net red velocity vector oscillates around
    the solid black vector and produces a conical surface with $\psi$
    as its half-opening angle.  Note that $v_\mathrm{orb}$ is
    exagerated here.  In practice, $v_\mathrm{orb}\ll
    v_\mathrm{jet}\sim c$.  This figure then shows that
    $\sin{\psi}\sim\tan{\psi}=v_\mathrm{orb}\cos{\chi}/v_\mathrm{jet}$.}
  \label{fig:jet}
\end{figure}

In the GW dominated phase of binary's evolution, recent simulations
show that the BHs can continue to accrete gas from the circumbinary
disk all the way until merger \citep{2015MNRAS.446L..36F}.  At this
stage, we consider a ballistic jet emanating from one of binary's
member BHs at speed $\beta=v_\mathrm{jet}/c$ at an angle $\chi$ to the
direction of the orbital angular momentum.  This geometry is depicted
in Figures~\ref{fig:BH_orbit} and \ref{fig:jet}.  As a result of the
binary orbit, the jet will wiggle on a conical surface.  (In the limit
of a small binary velocity, i.e., at large relative separations, the
jet material will move on a cylindrical surface.)  The half-opening
angle of the cone is given by \citep{1993ApJ...409..130R}
\begin{equation}
  \sin\psi = \frac{v_\mathrm{orb}\cos\chi}{v_\mathrm{jet}}.
\end{equation}
When the binary separation is where gravitational wave-induced
inspiral begins, it has a velocity given by equation~(\ref{eqn:v}),
and the separation between the two black holes at this instant is
given by equation~(\ref{eqn:a}).  Also, from equation
(\ref{eqn:vorb}), the orbital velocity of each BH evolves as
\begin{equation}
  \frac{\dot v_\mathrm{orb}}{v_\mathrm{orb}} = -\frac{1}{2}\frac{\dot a}{a}.
\end{equation}
which gives the evolution of the opening angle of the cone as 
\begin{equation}
  \dot\psi^\mathrm{obs}=\dot\psi = -\frac{1}{2\cot\psi}\frac{\dot a}{a}.
\end{equation}
This can be solved analytically to get
\begin{equation}
  \sin\psi = \sin\psi_0\left(\frac{a_0}{a}\right),
\end{equation}
where $\psi_0$ and $a_0$ are the initial values. Thus, as the binary
continues to shrink the BH orbital velocity grows and the opening
angle of the conical jet increases (cone opens wider).  The rate at
which this happens reflects the binary's dynamical evolution.
Figure~\ref{fig:evolve} shows the evolution of the half-opening angle
in the gas and GW-dominated phase.

@@

\begin{figure*}
\begin{center}
  \begin{tabular}{cc}
    % jet3d_check2.py (case 0) 
    \includegraphics*[scale=0.45]{{jet_i40_beta0.90_mdot1.00_full}.pdf} &
    % jet3d_image_fulljet.py (case 1) 
    \includegraphics*[scale=0.45]{jet_check_zoom1.pdf}
  \end{tabular}
\end{center}
\caption{The left panel shows forward and backward jets for an
  equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  We have
  assumed $i=40^\circ$, $\beta=0.9$, and $d = 100$~Mpc.  The jets are
  twisted because of the apparant superluminal motion.  Also evident
  is the opening of the jet and the radio jet chirp.  Right panel
  shows normalised brightness of the inner 30~mas region of the
  forward jet.}
\label{fig:jet_mdot1.0}
\end{figure*}

We now ask if it would be possible to observe the opening of the jet
morphology.  In order to do this, it is necessary to map the above
calculations to the observer's frame \citep{1982ApJ...262..478G}.  At
a large distance compared to the binary's separation, the jet geometry
is illustrated in Figure \ref{fig:jetgeom}.  Here the source frame of
reference is shown by $x^\prime$, $y^\prime$, $z^\prime$ coordinate
systems.  The $x$, $y$, $z$ coordinate system represents the
observer's frame such that the $yz$ is the plane of the sky.  Further,
the jet is oriented such that that the $y^\prime$ axis coincides with
the $y$ axis, the original direction of jet emission is at an angle
$i$ relative to the line of sight ($x$-axis).  For simplicity, we also
assume that the $z^\prime$ axis is in the $xz$ plane.  Then the jet
will precess on a cone with a half-opening angle $\psi$ and an angular
velocity of precession $\Omega$.  In the source frame of reference,
the equation of the jet is then given by \citep{1981ApJ...246L.141H}
\begin{eqnarray}
  &v_x=&s_\mathrm{jet}\beta c\left[\sin\psi\sin i\cos\Omega t+\cos\psi\cos i\right],\\
  &v_y=&s_\mathrm{jet}\beta c\sin\psi\sin\Omega t,~\,\mathrm{and}\\
  &v_z=&s_\mathrm{jet}\beta c\left[\cos\psi\sin i-\sin\psi\cos i\cos\Omega t\right]. 
\end{eqnarray}
Here $s_\mathrm{jet}=\pm 1$ for forward and backward jets,
respectively.  The coordinates of a jet particle in the sky is then
given by $z=v_zt$ and $y=v_yt$.  However, because of superluminal
motion, the observer will see this on the sky as
\begin{equation}
  z=v_zt/(1-v_x/c)
\end{equation}
and
\begin{equation}
  y=v_yt/(1-v_x/c)
\end{equation}
respectively.  In angular coordinates, the motion on the sky is given
by $\phi_z=z/d$ and $\phi_y=y/d$ where $d$ is the angular diameter
distance of the binary.  Because of superluminal motion, the forward
jet is stretched while the backward jet is compressed.

Finally, we assume that the emission from each jet particle is
optically thin at the frequency of observation with a spectrum in the
rest frame of the particle given by $P(\nu)\propto \nu^{-\alpha}$
\citep{1982ApJ...262..478G}.  The observed flux density $S(\nu)$ in erg
cm$^{-2}$ Hz$^{-1}$ is given by
\begin{equation}
  S(\nu)=S_r(\nu)D^{3+\alpha},
\end{equation}
where $D$ is the Doppler shift parameter
$[\gamma(1-\beta\cos\phi)]^{-1}$ and $S_r$ is the rest frame flux
density.  Further to consider the effects of jet evolution we assume
that $S_r$ decreases with time according to a simple power law
\begin{equation}
  S_r\propto t^{-\delta}.
\end{equation}
We assume $\delta=1$ and $\alpha=1$, hereafter
\citep{1982ApJ...262..478G}.

@@

As the binary evolves, its increasing orbital period and speed, and
the apparant superluminal motion effect, together result in a jet
morphology with an increasing opening angle and increasingly rapid
periodic variation, together with a twisted shape.
Figure~\ref{fig:jet_mdot1.0} shows these three features for an
equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  For
concreteness, we have assumed $i=40^\circ$ and $\beta=0.9$.  The
angular diameter distance to the binary is assumed to be $d =
100$~Mpc, which corresponds to redshift $z\sim 0.3$.  The forward jet
streched on the sky while the backward jet is compressed due to the
superluminal motion.  The right panel of Figure~\ref{fig:jet_mdot1.0}
zooms into the inner region of the forward jet and shows that the
binary creates wiggles in the jet morphology of the order of a
milli-arcsecond.  The duration of the binary evolution captured by
Figure~\ref{fig:jet_mdot1.0} is 100 yr.  Although the system can be
rescaled to different masses, times, and separation, the initial
separation in Figure~\ref{fig:jet_mdot1.0} is $\sim 10^{-3}$~pc, when
the binary enters it GW-dominated phase.

The increasing orbital speed of the binary leads to an increase in the
jet's precession angular velocity.  This increases the helicity of the
jet such that the jet is wound up increasingly tightly closer to the
source.  This gives rise to a ``chirping'' morphology as seen in
Figure~\ref{fig:jet_mdot1.0}.  The chirp is easier to resolve in jets
that are closer to the line of sight, where superluminal motion can
magnify the forward jet.  A chirping black hole is an unambiguous
signature of a rapidly evolving binary BH; the chirp is direct
reflection of the orbital velocity evolution in
Figure~\ref{fig:evolve}.  Together, the three features of the jet
morphology in Figure~\ref{fig:jet_mdot1.0}, viz., conical jet geometry
with an increasing opening angle, a radio chirp geometry, and an
asymmetrical shape, directly probe the dynamics of the binary in its
GW phase. Observations of this milliarcsecond geometry can thus
constrain GR in the strong field limit.

@@

Using a model of jet precession in a binary BH, we have shown that the
jet morphology of a binary that is entering its GW-induced phase of
inspiral shows a distinct chirping morphology. Jets in these binaries
have a biconical morphology.  After the binary enters the GW phase,
the opening angle of cone rapidly increases within a distance of tens
of milli-arcseconds.  In addition, the periodicity of the jet's motion
on the conical surface increases as the orbital period of the binary
goes up in this stage of its evolution.  Finally, due to apparent
superluminal motion and the widening opening angle, the jet is twisted
into a peculiar morphology.  The jet geometry can be used to infer the
set of dynamical and geometrical parameters, listed in
Table~\ref{table:params}.

>>>>

\begin{abstract}
  We study a novel electromagnetic signature of supermassive black
  hole binaries whose inspiral starts being dominated by gravitational
  wave (GW) emission. Combined GW and electromagnetic observations of
  compact binary mergers would allow to use them as standard sirens
  for cosmology and a testing ground of general relativity.  If one of
  the binary members produces a radio jet, the helical chirping
  morphology of the jet can be used to infer the binary parameters.
  The jet precesses along a biconical surface due to the binary's
  orbital motion.  When the binary enters the GW phase of its
  evolution, the opening angle widens, the jet exhibits milliarcsecond
  scale wiggles and the conical surface of jet precession is twisted
  due to the superluminal motion.  For binaries with mass
  $10^7$--$10^{10}$ M$_\odot$ at redshifts $z<0.5$, monitoring these
  features in current and archival data will place a lower limit on
  sources that could be detected by eLISA and Pulsar Timing Arrays.
  In the future, microarcsecond interferometry with the Square
  Kilometer Array will increase the potential usefulness of this
  technique.
\end{abstract}

@@

Binaries of supermassive black holes (BHs) arise naturally as a result
of mergers of galaxies in the context of the heirarchical galaxy
formation \citep{2000MNRAS.311..576K, 2002MNRAS.336L..61H,
  2008ApJ...676...33D, 2012MNRAS.422.1306K}.  After a galaxy merger,
infalling black holes lose their angular momentum in three stages
before coalescence \citep{1980Natur.287..307B, 2005LRR.....8....8M,
  2011ASL.....4..181C}.  In the first stage, the black holes sink to
the center of the gravitational potential of the merger remnant due to
dynamical friction and form a gravitationally bound binary.  The newly
formed binary then continues to lose its angular momentum and shrink
by scattering ambient gas and stars.  In the final stage of the
binary's evolution, emission of gravitational waves (GWs) becomes the
predominant mode of angular momentum loss, which results in the
coalescence of the black holes.

General relativisitic (GR) numerical simulations can predict the
binary's evolution in its final stage up to coalescence
\citep{2005PhRvL..95l1101P, 2006PhRvL..96k1102B, 2006PhRvL..96k1101C}.
This has stimulated interest in predicting related electromagnetic
signals of black hole coalescence.  Effects considered in the
literature include afterglows due to infall of gas on the remnant
black hole \citep{2005ApJ...622L..93M}, electromagnetic variability in
the circumbinary disk due to shocks induced by the sudden mass loss of
the binary following GW emission at coalescence
\citep{2007APS..APR.S1010B}, flares from shocked remnants of accretion
disks around recoiled remnant black hole \citep{2007PhRvL..99d1103L,
  2008ApJ...682..758S, 2009CQGra..26i4032H}, quasi-periodic
variability in gas accretion in the early stages of black hole merger
\citep{2008ApJ...672...83M, 2009CQGra..26i4032H, 2012MNRAS.427.2680K},
evolution in the profiles of broad emission lines shortly before
merger \citep{2013MNRAS.432.1468M}, characteristic evolution in
thermal emission due to circumbinary accretion
\citep{2015MNRAS.446L..36F}, brightening of accretion disk due to
viscous dissipation of gravitational waves
\citep{2008PhRvL.101d1101K}, and prompt tidal disruption of stars by
the recoiled remnant \citep{2011MNRAS.412...75S}.

In addition to testing GR in the strong field limit, observations of
electromagnetic counterparts will help localize sources of GWs for
future detectors such as the Evolved Laser Interferometer Space
Antenna (eLISA)\footnote{\url{https://www.elisascience.org}}, which
would be capable of observing the peak GW luminosity for BH
coalescence of $\sim 10^{57}$ erg$/$s out to cosmological distances
\citep{2003CQGra..20S..65H, 2013CQGra..30x4009S}.  Simultaneous
detection of GW and electromagnetic signals from coalescing binaries
will yield rich scientific pay-offs \citep{2003CQGra..20S..65H,
  2005ApJ...629...15H}.  For example, redshift measurement of eLISA
sources could constraint cosmological parameters.  Measurement of the
black hole masses and spins from the GW signal could constrain
accretion physics during coalescence.  But even before eLISA is
launched, a search of electromagnetic signatures may help calibrate
the expected abundance of GW sources.

In this Letter we study the morphology of a jet emitted by one of the
binary members during its final phase of evolution before coalescence,
i.e., when the binary's dynamics is dominated by GW emission.  We show
that the helical chirping morphology of the jet can be used to infer
the binary parameters.

@@

We study a BH binary formed following a gas-rich merger of two
galaxies as a result of the processes described above.  We consider
black holes with masses $M_1$ and $M_2$ in a circular orbit of radius
$a$ around each other.  Let $M=M_1+M_2$.  For simplicity, we consider
an equal-mass binary ($M_1=M_2$) on a circular orbit. The orbital
speed of each black hole is given by \citep{2010PhRvD..81d7503L}
\begin{equation}
  v = \frac{1}{2}\left(\frac{GM}{a}\right)^{1/2}\!\!\!\! = 5.8\times 10^3\, \mathrm{km~s}^{-1}M_8^{1/2}a_{16}^{-1/2},
  \label{eqn:vorb}
\end{equation}
where $a_{16}\equiv (a/10^{16}$ cm$)$ and $M_8\equiv (M/10^8$
M$_\odot)$ are the binary separation and mass,
respectively.\footnote{Note that $10^{16} \mathrm{cm}\approx 3.4\times
  10^{-3} \mathrm{pc}$.  The Schwarzchild radius of a $10^8$~M$_\odot$
  black hole is $9.6\times 10^{-6}$~pc, or $\sim 3\times 10^{13}$~cm.}
The orbital period is
\begin{equation}
  P = 2\pi\left(\frac{GM}{a^3}\right)^{-1/2}\!\!\!\! = 1.72\, \mathrm{yr}\, a_{16}^{3/2} M_8^{-1/2}.
\end{equation}

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \draw (0,0) ellipse (3 and 0.5);
      \fill[black] (0,0) circle (1pt);
      \draw [-angle 45,thick] (2,-0.39) -- (3.5,-0.1);
      \draw [-angle 45,thick,dashed,dash pattern = on 4pt off 1pt] (2,-0.39) -- (2,1.5);
      \draw [-angle 45,thick] (2,-0.39) -- (2.5,0.8);
      \draw [-angle 45,thick,red] (2,-0.39) -- (3.5,0.8);
      \fill[black] (-2,0.39) circle (2pt);
      \fill[black] (2,-0.39) circle (2pt);
      \draw (4.1,-0.1) node {\large ${\mathbf{v}}_\mathrm{orb}$};
      \draw (2.7,1.0) node {\large ${\mathbf{v}}_\mathrm{jet}$};
      \draw (4.0,1.0) node {\large $\tilde{\mathbf{v}}_\mathrm{jet}$};
      \draw (2.0,1.8) node {\large $\mathbf{L}_\mathrm{orb}$};
    \end{tikzpicture}
  \end{center}
  \caption{Jet and orbital configuration. The orbital velocity
    $\mathbf{v}_\mathrm{orb}$ causes the jet to precess as its source
    BH moves along its binary orbit. The net jet velocity is
    $\tilde{\mathbf{v}}_\mathrm{jet}={\mathbf{v}}_\mathrm{orb}+{\mathbf{v}}_\mathrm{jet}$,
    where $\mathbf{v}_\mathrm{jet}$ is its ejection velocity of the
    jet material at the source.}
  \label{fig:BH_orbit}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \draw [-angle 45,dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (3.0,0);
      \draw [-angle 45,dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (0,4.0);
      \draw [-angle 45,thick] (0,0) -- (2.0,4.0);
      \draw [-angle 45,thick,red] (0,0) -- (5.0,4.0);
      \draw (1.0,2.0) arc (63:90:2.2);
      \draw (2.0,1.6) arc (38.6:63:2.56);
      \draw (0.5,2.5) node {\large $\chi$};
      \draw (1.90,2.3) node {\large $\psi$};
      \draw [dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (3.0,6.0);
      \draw (0.0,4.5) node {\large $\mathbf{L}_\mathrm{orb}$};
      \draw (3.5,0.0) node {\large $\mathbf{v}_\mathrm{orb}$};
      \draw (1.0,3.0) node {\large $\mathbf{v}_\mathrm{jet}$};
      \draw (5.3,4.2) node {\large \textcolor{red}{$\mathbf{\tilde v}_\mathrm{jet}$}};
    \end{tikzpicture}
  \end{center}
  \caption{The orbital velocity $\mathbf{v}_\mathrm{orbital}$
    introduces the jet precision angle $\psi$.  Without the orbital
    motion, $\psi=0$, and the jet traces a cylindrical surface.  With
    the orbital motion, the net red velocity vector oscillates around
    the solid black vector and produces a conical surface with $\psi$
    as its half-opening angle.  Note that $v_\mathrm{orb}$ is
    exagerated here.  In practice, $v_\mathrm{orb}\ll
    v_\mathrm{jet}\sim c$.  The figure indicates that
    $\sin{\psi}\sim\tan{\psi}=v_\mathrm{orb}\cos{\chi}/v_\mathrm{jet}$.}
  \label{fig:jet}
\end{figure}

In the GW dominated phase of binary's evolution, recent simulations
show that the BHs can continue to accrete gas from the circumbinary
disk all the way until merger \citep{2015MNRAS.446L..36F}.  At this
stage, we consider a ballistic jet emanating from one of binary's
member BHs at speed $\beta=v_\mathrm{jet}/c$ at an angle $\chi$ to the
direction of the orbital angular momentum.  This geometry is depicted
in Figures~\ref{fig:BH_orbit} and \ref{fig:jet}.  As a result of the
binary orbit, the jet will wiggle on a conical surface.  (In the limit
of a small binary velocity, i.e., at large relative separations, the
jet material will move on a cylindrical surface.)  The half-opening
angle of the cone is given by \citep{1993ApJ...409..130R}
\begin{equation}
  \sin\psi = \frac{v_\mathrm{orb}\cos\chi}{v_\mathrm{jet}}.
\end{equation}
When the binary separation reaches the point where gravitational
wave-induced inspiral begins, the binary has a velocity given by
equation~(\ref{eqn:v}), and a separation given by
equation~(\ref{eqn:a}).  From equation (\ref{eqn:vorb}), the orbital
velocity of each BH evolves as
\begin{equation}
  \frac{\dot v_\mathrm{orb}}{v_\mathrm{orb}} = -\frac{1}{2}\frac{\dot a}{a}.
\end{equation}
which gives the evolution of the opening angle of the cone as 
\begin{equation}
  \dot\psi^\mathrm{obs}=\dot\psi = -\frac{1}{2\cot\psi}\frac{\dot a}{a}.
\end{equation}
This can be solved analytically to get
\begin{equation}
  \sin\psi = \sin\psi_0\left(\frac{a_0}{a}\right),
\end{equation}
where $\psi_0$ and $a_0$ are the initial values. Thus, as the binary
continues to shrink the BH orbital velocity grows and the opening
angle of the conical jet increases (cone opens wider).  The rate at
which this happens reflects the binary's dynamical evolution.
Figure~\ref{fig:evolve} shows the evolution of the half-opening angle
in the gas and GW-dominated phase.

@@

\begin{figure*}
\begin{center}
  \begin{tabular}{cc}
    % jet3d_check2.py (case 0) 
    \includegraphics*[scale=0.45]{{jet_i40_beta0.90_mdot1.00_full}.pdf} &
    % jet3d_image_fulljet.py (case 1) 
    \includegraphics*[scale=0.45]{jet_check_zoom1.pdf}
  \end{tabular}
\end{center}
\caption{The left panel shows forward and backward jets for an
  equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  We have
  assumed $i=40^\circ$, $\beta=0.9$, and angular diameter distance $d
  = 100$~Mpc.  The jets are twisted because of the apparant
  superluminal motion.  Also evident is the opening of the jet and the
  radio jet chirp.  Right panel shows normalised brightness of the
  inner 30~mas region of the forward jet.}
\label{fig:jet_mdot1.0}
\end{figure*}

We now ask if it would be possible to observe the opening of the jet
morphology.  For this purpose, it is necessary to map the above
calculations to the observer's frame \citep{1982ApJ...262..478G}.  At
a large distance compared to the binary's separation, the jet geometry
is illustrated in Figure \ref{fig:jetgeom}.  Here the source frame of
reference is shown by $x^\prime$, $y^\prime$, $z^\prime$ coordinate
systems.  The $x$, $y$, $z$ coordinate system represents the
observer's frame such that the $yz$ is the plane of the sky.  Further,
the jet is oriented such that the $y^\prime$ axis coincides with the
$y$ axis, the original direction of jet emission is at an angle $i$
relative to the line of sight ($x$-axis).  For simplicity, we also
assume that the $z^\prime$ axis is in the $xz$ plane.  Then the jet
will precess on a cone with a half-opening angle $\psi$ and an angular
velocity of precession $\Omega$.  In the source frame of reference,
the equation of the jet is then given by \citep{1981ApJ...246L.141H}
\begin{eqnarray}
  &v_x=&s_\mathrm{jet}\beta c\left[\sin\psi\sin i\cos\Omega t+\cos\psi\cos i\right],\\
  &v_y=&s_\mathrm{jet}\beta c\sin\psi\sin\Omega t,~\,\mathrm{and}\\
  &v_z=&s_\mathrm{jet}\beta c\left[\cos\psi\sin i-\sin\psi\cos i\cos\Omega t\right]. 
\end{eqnarray}
Here $s_\mathrm{jet}=\pm 1$ for forward and backward jets,
respectively.  The coordinates of a jet particle in the sky is then
given by $z=v_zt$ and $y=v_yt$.  However, because of superluminal
motion, the observer will see this on the sky as
\begin{equation}
  z=v_zt/(1-v_x/c)
\end{equation}
and
\begin{equation}
  y=v_yt/(1-v_x/c)
\end{equation}
respectively.  In angular coordinates, the motion of the jet on the
sky is given by $\phi_z=z/d$ and $\phi_y=y/d$ where $d$ is the angular
diameter distance of the binary.  Because of superluminal motion, the
forward jet is stretched while the backward jet is compressed.

Finally, we assume that the emission from each jet is optically thin
at the frequency of observation with a spectrum in the rest frame of
the particle given by $P(\nu)\propto \nu^{-\alpha}$
\citep{1982ApJ...262..478G}.  The observed flux density $S(\nu)$ in
erg cm$^{-2}$ Hz$^{-1}$ is given by
\begin{equation}
  S(\nu)=S_r(\nu)D^{3+\alpha},
\end{equation}
where $D$ is the Doppler shift parameter
$[\gamma(1-\beta\cos\phi)]^{-1}$ and $S_r$ is the rest frame flux
density.  Further, to consider the effects of jet evolution we assume
that $S_r$ decreases with time according to a simple power law
\begin{equation}
  S_r\propto t^{-\delta}.
\end{equation}
We assume $\delta=1$ and $\alpha=1$, hereafter
\citep{1982ApJ...262..478G}.

@@

As the binary evolves, the combination of increasing orbital period
and speed, and the apparant superluminal motion effect, results in a
jet morphology with an increasing opening angle and increasingly rapid
periodic variation, together with a twisted shape.
Figure~\ref{fig:jet_mdot1.0} shows these three features for an
equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  For
concreteness, we have assumed $i=40^\circ$ and $\beta=0.9$.  The
angular diameter distance to the binary is assumed to be $d =
100$~Mpc, which corresponds to redshift $z\sim 0.3$.  The forward jet
streches on the sky while the backward jet is compressed due to the
superluminal motion.  The right panel of Figure~\ref{fig:jet_mdot1.0}
zooms into the inner region of the forward jet and shows that the
binary creates wiggles in the jet morphology of the order of a
milli-arcsecond.  The duration of the binary evolution captured by
Figure~\ref{fig:jet_mdot1.0} is 100 yr.  Although the system can be
rescaled to different masses, times, and separation, the initial
separation in Figure~\ref{fig:jet_mdot1.0} is $\sim 10^{-3}$~pc, when
the binary enters it GW-dominated phase.

The increasing orbital speed of the binary leads to an increase in the
jet's precession angular velocity.  This increases the helicity of the
jet such that the jet is wound up increasingly tightly closer to the
source.  This gives rise to a ``chirping'' morphology as seen in
Figure~\ref{fig:jet_mdot1.0}.  The chirp is easier to resolve in jets
that are closer to the line of sight, where superluminal motion can
magnify the forward jet.  A chirping black hole is an unambiguous
signature of a rapidly evolving binary BH system; the chirp is direct
reflection of the orbital velocity evolution in
Figure~\ref{fig:evolve}.  Together, the three features of the jet
morphology in Figure~\ref{fig:jet_mdot1.0}, namely, conical jet
geometry with an increasing opening angle, a radio chirp geometry, and
an asymmetrical shape, directly probe the dynamics of the binary in
its GW phase. Observations of such a milliarcsecond geometry can thus
constrain GR in the strong field limit.

@@

Using a model of jet precession in a binary BH, we have shown that the
jet morphology of a binary that is entering its GW-induced phase of
inspiral exhibits a distinct chirping morphology. Jets in these
binaries have a biconical morphology.  After the binary enters the GW
phase, the opening angle of cone rapidly increases within a separation
of tens of milli-arcseconds at a distance of 100~Mpc.  In addition,
the periodicity of the jet's motion on the conical surface increases
as the orbital period of the binary goes up in this stage of its
evolution.  Finally, due to the apparent superluminal motion and the
widening opening angle, the jet is twisted into a peculiar morphology.
The jet geometry can be used to infer the set of dynamical and
geometrical parameters, listed in Table~\ref{table:params}.

Milli-arcsecond features can be resolved using long baseline radio
interferometry \citep{2001ChJAA...1..236Q}.  At higher redshifts, jets
are fainter and the wiggles become more difficult to resolve.
However, it might be possible to resolve binaries with very unequal
mass ratios, eccentric orbits, or suitable inclinations.  The future
Square Kilometer Array (SKA) should make microarcsecond resolution
possible, which could increase the detection rate of these objects
\citep{2014arXiv1412.5971P}.
